Classless Java

### 2015-11-22 paper/sections/CaseStudy.tex
EP case study text revised with independent extensibility

The \emph{Expression Problem} (EP)~\cite{wadler98expression-problem} is coined
by Wadler about modular extensibility issues in software evolution and has been
a hot topic in programming languages since. Today we know of various solutions
to the EP that either rely on \emph{new programming language
  features}~\cite{chambers95multimethods,clifton00multiJava,madsen89virtual,nystrom06j,bruce98astatically,McDirmid01jiazzi,garrigue98polymorphic,zenger01extensible,loh06open,wehr11javaGI},
or can be used as \emph{design patterns}~\cite{gof} in existing
languages~\cite{torgersen04theexpression,oliveira06extensible,wouter08datatypes,oliveira09modular,oliveira12extensibility}. In
this paper we show that the EP can be solved with our \mixin annotation in a
trivial way. We will use a canonical example of implementing arithmetic
expressions gradually to illustrate the usage.

\subsubsection{Initial System}
The initial system expresses arithmetic expressions with only literals and
addition structure:

@@

\subsubsection{Adding a New Variant}
It is easy to add new data variants to the code in the initial system in a
modular way. For example, the following code shows how to add the subtraction variant.

@@

\subsubsection{Adding a New Operation}
Adding new operations to the previous system is still straightforward (although
requires longer code). The following code shows an example of adding a new
operation \texttt{print}.

@@

>>>>

The \emph{Expression Problem} (EP)~\cite{wadler98expression-problem} is coined
by Wadler about modular extensibility issues in software evolution and has been
a hot topic in programming languages since. The requirements of a solution to
the EP can be summarized as: 
\begin{itemize}
\item \emph{Extensibility in both dimensions}: A solution must allow
  the addition of new data variants and new operations and support
  extending existing operations.
\item \emph{Strong static type safety}: A solution must prevent
  applying an operation to a data variant which it cannot handle using
  static checks.
\item \emph{No modification or duplication}: Existing code must not be
  modified nor duplicated.
\item \emph{Separate compilation and type-checking}: Safety checks or
  compilation steps must not be deferred until link or runtime.
\item \emph{Independent extensibility}~\cite{zenger05independentlyextensible}:
  It should be possible to combine independently developed extensions so that
  they can be used jointly.
\end{itemize}

Today we know of various solutions to the EP that either rely on \emph{new
  programming language features}
% ~\cite{chambers95multimethods, clifton00multiJava, madsen89virtual, nystrom06j,
%   bruce98astatically, McDirmid01jiazzi, garrigue98polymorphic,
%   zenger01extensible, loh06open, wehr11javaGI}, 
or can be used as \emph{design patterns}%~\cite{gof}
in existing languages.
% ~\cite{torgersen04theexpression, oliveira06extensible,
%  wouter08datatypes, oliveira09modular, oliveira12extensibility}. 
Now we show that the EP can be solved with our \mixin annotation in a trivial
way. We will use the canonical example of implementing arithmetic expressions
gradually to illustrate the usage.

\paragraph{Initial System}
The initial system expresses arithmetic expressions with only literals and
addition structure:

@@

\paragraph{Adding a New Variant}
It is easy to add new data variants to the code in the initial system in a
modular way. For example, the following code shows how to add the subtraction
variant.

@@

\paragraph{Adding a New Operation}
Adding new operations to the previous system is still straightforward (although
requires longer code). The following code shows an example of adding a new
operation \texttt{print}.

@@

\paragraph{Independent Extensibility}
To show that our approach supports independent extensibility, we first define a
new operation \texttt{collectLit}(which collects all
literal components in an expression) on expressions:

\lstinputlisting[linerange=70-86]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=EXPRESSION_COLLECTLIT

\noindent Now we combine the two extensions (\texttt{print} and
\texttt{collectLit}) together:

\lstinputlisting[linerange=90-94]{../UseMixinLombok/src/test/TestExpression.java}% APPLY:linerange=INDEPENDENT_EXTENSIBILITY

\noindent \texttt{ExpPC} is the new expression interface supporting
\texttt{print} and \texttt{collectLit} operations; \texttt{LitPC} and
\texttt{AddPC} are the extended variants. Notice that except for the routine of
\textbf{extends} clauses, no glue code is required. Return types of
\texttt{e1,e2} are also automatically refined to \texttt{ExpPC}.
